---
title: "Covid urine data analysis with Human GEM"
author: "Ziwei Huang"
date:   "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: united
    code_folding: hide
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = FALSE,warning = FALSE)
```

## Setting
```{r}
rm(list=ls())
suppressPackageStartupMessages(library(Biobase))
suppressPackageStartupMessages(library(GSVA))
suppressPackageStartupMessages(library(Matrix))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lme4))
suppressPackageStartupMessages(library(lmerTest))
suppressPackageStartupMessages(library(hypeR))
suppressPackageStartupMessages(library(gprofiler2))
suppressPackageStartupMessages(library(DT))


PATH <- file.path(Sys.getenv("NECS"),"Ziwei_Analysis","hypeR.GEM_evaluation")
devtools::load_all(file.path(PATH, "../in_dev_packages/hypeR.GEM"))

source(file.path(PATH,"R/utility.R"))
```

## Load metabolite data and convert to refmet_name {.tabset}
```{r}
do.run <- FALSE

if(do.run){
  max_fdr <- 0.05
  
  met_urine <- read_excel(file.path(PATH,"/data/External/covid_S2_matrix.xlsx"),
                   sheet = "Urine_metabolite_matrix_raw",
                   skip = 0) %>% 
    dplyr::rename(metabolite = "Metabolite") %>% 
    as.data.frame(.)
  
  ## Differential analysis
  met_urine_DE <- read_excel(file.path(PATH,"/data/External/covid_S4_DE.xlsx"),
                   sheet = "DEMs_urine",
                   skip = 0) %>% 
    dplyr::rename(metabolite = "Metabolite",
                  estimate = "Log2（fold change）") %>% 
    as.data.frame(.)
  
  ## convert to refmet_name
  met_urine_refmet <- refmet_convert(DF=met_urine)
  
  met_urine_DE <- met_urine_DE %>% 
    dplyr::left_join(met_urine_refmet, by="metabolite") %>% 
    dplyr::select(metabolite, refmet_name, estimate, P_value_adjust) %>% 
    dplyr::mutate(gene_type  = case_when(
        estimate > 0 & P_value_adjust < max_fdr ~ "up",
        estimate < 0 & P_value_adjust < max_fdr ~ "dn",
        TRUE ~ "ns"
      ))
  ## create signature list
  met_urine_sigs <- list(up = met_urine_DE %>% dplyr::filter(gene_type == "up"),
                         dn = met_urine_DE %>% dplyr::filter(gene_type == "dn"))
  saveRDS(met_urine_sigs, file = file.path(PATH,"/data/External/COVID_urine_mets.rds"))
}else{
  met_urine_sigs <- readRDS(file = file.path(PATH,"/data/External/COVID_urine_mets.rds"))
}

## check
sapply(met_urine_sigs ,nrow) %>%  tibble::enframe(value = "size")
```

## Load protein data {.tabset}
```{r}
if(do.run){
  max_fdr <- 0.01
  
  prot_urine <- read_excel(file.path(PATH,"/data/External/covid_S2_matrix.xlsx"),
                   sheet = "Urine_protein_matrix",
                   skip = 0) %>% 
    tidyr::separate_wider_delim(Protein,
                                 "_",
                                 names = c("tag", "symbol")) %>% 
    dplyr::select(-c("tag"))
  
  ## Differential analysis
  
  prot_urine_DE <- read_excel(file.path(PATH,"/data/External/covid_S4_DE.xlsx"),
                   sheet = "DEPs_urine",
                   skip = 0) %>% 
    dplyr::rename(estimate = "Log2（fold change）") %>% 
    tidyr::separate_wider_delim(Protein,
                                 "_",
                                 names = c("tag", "symbol")) %>% 
    dplyr::select(-c("tag")) %>% 
    dplyr::mutate(gene_type  = case_when(
        estimate > 0 & P_value_adjust < max_fdr ~ "up",
        estimate < 0 & P_value_adjust < max_fdr ~ "dn",
        TRUE ~ "ns"
      ))
  
  prot_urine <- prot_urine %>% 
    dplyr::select(symbol) %>% 
    dplyr::left_join(prot_urine_DE,by='symbol') %>% 
    dplyr::mutate(gene_type  = case_when(
        gene_type == "up" ~ "up",
        gene_type == "dn" ~ "dn",
        TRUE ~ "ns"
      ))
 
  saveRDS(prot_urine, file = file.path(PATH,"/data/External/COVID_urine_protein.rds"))
  saveRDS(prot_urine$symbol, file = file.path(PATH,"/data/External/all_protein_urine.rds"))
}else{
  protein_diffanal <- readRDS(file = file.path(PATH,"/data/External/COVID_urine_protein.rds"))
  all_prot_urine <- readRDS(file = file.path(PATH,"/data/External/all_protein_urine.rds"))
}

prot_urine_sigs <- list(up = protein_diffanal %>% dplyr::filter(gene_type == "up"),
                        dn = protein_diffanal %>% dplyr::filter(gene_type == "dn"))

## check
sapply(prot_urine_sigs ,nrow) %>%  tibble::enframe(value = "size")
```

## GEM mapping {.tabset}

```{r}
data("Human_GEM_tables")
GEM_gene_map <- gprofiler2::gconvert(query = Human_GEM_tables$gene_df$name,
                                      organism = "hsapiens",
                                      target = "ENSG",
                                      filter_na = TRUE) %>%
   dplyr::select(input, name) %>%
   dplyr::rename(ensemble_ID = input, symbol = name)

```

### Undirectional
```{r}
hypeR_GEM_obj <- hypeR.GEM::signature2gene(signatures = met_urine_sigs,
                                           species = "Human",
                                           directional = FALSE,
                                           merge = TRUE,
                                           promiscuous_threshold = 500,
                                           ensemble_id = TRUE,
                                           reference_key = 'refmet_name',
                                           background = NULL)
## check
sapply(hypeR_GEM_obj$gene_tables,nrow) %>%  tibble::enframe(value = "size")

saveRDS(hypeR_GEM_obj, file = file.path(PATH, "data/External/urine_hypR_object.rds"))
```


### Directional
```{r}
hypeR_GEM_obj_di <- hypeR.GEM::signature2gene(signatures = met_urine_sigs,
                                           species = "Human",
                                           directional = TRUE,
                                           merge = TRUE,
                                           promiscuous_threshold = 500,
                                           ensemble_id = TRUE,
                                           reference_key = 'refmet_name',
                                           background = NULL)
## check
sapply(hypeR_GEM_obj_di$gene_tables,nrow) %>%  tibble::enframe(value = "size")

saveRDS(hypeR_GEM_obj_di, file = file.path(PATH, "data/External/urine_hypR_object_di.rds"))
```

## Pair evaluation (gene-level, hypergeometric) {.tabset}

- Take overlap between hypeR-GEM mapped genes and genes in proteomics ($GEM \cap genes\ in\ proteomics$)

- Counter how many of these overlap genes are significant

- Background = ($total GEM \cap proteomics$)

- setA = ($GEM\ mapped \cap Background$)

- setB = ($significant\ genes\ in\ proteomics \cap Background$)

```{r}
tmp <- sapply(met_urine_sigs ,nrow) %>%  tibble::enframe(value = "size")

background_list <- intersect(GEM_gene_map$symbol,all_prot_urine)


table_protein <- pair_evaluation(hypeR_GEM_obj,
                                protein_diffanal,
                                background_list) 

table_protein$df_grouped <- table_protein$df_grouped %>% 
  dplyr::mutate(DE_met = sum(tmp$size)) %>% 
  dplyr::select(DE_met, everything())

table_protein_di <- pair_evaluation(hypeR_GEM_obj_di,
                                  protein_diffanal,
                                  background_list) 

table_protein_di$df_grouped <- table_protein_di$df_grouped %>% 
  dplyr::mutate(DE_met = sum(tmp$size)) %>% 
  dplyr::select(DE_met, everything())


saveRDS(table_protein, file = file.path(PATH, "data/External/urine_table_protein.rds"))
saveRDS(table_protein_di, file = file.path(PATH, "data/External/urine_table_protein_di.rds"))

```

### Undirectional vs. Protein_diffanal
```{r}
table_protein$df %>%  DT::datatable(.)

table_protein$df_grouped %>%  DT::datatable(.)
```


### Directional vs. Protein_diffanal
```{r}
table_protein_di$df %>%  DT::datatable(.)

table_protein_di$df_grouped %>%  DT::datatable(.)
```


## Pair evaluation (pathway-level, hypergeometric)

### Load genesets
```{r}
hallmark <- hypeR::msigdb_gsets(species="Homo sapiens", "H",clean = TRUE)
kegg <- hypeR::msigdb_gsets(species="Homo sapiens", "C2", "CP:KEGG_LEGACY", clean = TRUE)
reactome <- hyperdb_rgsets("REACTOME", "70.0")
gobp <- hypeR::msigdb_gsets(species="Homo sapiens", "C5", "BP", clean = TRUE)
gomf <- hypeR::msigdb_gsets(species="Homo sapiens", "C5", "MF", clean = TRUE)

## background of each omics
gene_background_GEM <- GEM_gene_map$symbol
gene_background_protein <- protein_diffanal$symbol
```

```{r}
hallmark_GEM <- reduce_gsets(hallmark, gene_background_GEM, min_size = 5)
cat("\n")
hallmark_protein <- reduce_gsets(hallmark, gene_background_protein, min_size = 5)

cat("\n")

hallmark_inter <- intersect(names(hallmark_GEM$genesets), names(hallmark_protein$genesets))
hallmark_GEM$genesets <- hallmark_GEM$genesets[hallmark_inter]
hallmark_protein$genesets <- hallmark_protein$genesets[hallmark_inter]

print(length(hallmark_inter))
```

```{r}
kegg_GEM <- reduce_gsets(kegg, gene_background_GEM, min_size = 5)
cat("\n")
kegg_protein <- reduce_gsets(kegg, gene_background_protein, min_size = 5)

cat("\n")

kegg_inter <- intersect(names(kegg_GEM$genesets), names(kegg_protein$genesets))
kegg_GEM$genesets <- kegg_GEM$genesets[kegg_inter]
kegg_protein$genesets <- kegg_protein$genesets[kegg_inter]

print(length(kegg_inter))
```


```{r}
reactome_GEM <- reduce_gsets(reactome, gene_background_GEM, min_size = 5)
cat("\n")
reactome_protein <- reduce_gsets(reactome, gene_background_protein, min_size = 5)

cat("\n")

reactome_inter <- intersect(names(reactome_GEM$genesets), names(reactome_protein$genesets))
reactome_GEM$genesets <- reactome_GEM$genesets[reactome_inter]
reactome_protein$genesets <- reactome_protein$genesets[reactome_inter]

print(length(reactome_inter))
```


```{r}
gobp_GEM <- reduce_gsets(gobp, gene_background_GEM, min_size = 5)
cat("\n")
gobp_protein <- reduce_gsets(gobp, gene_background_protein, min_size = 5)

cat("\n")

gobp_inter <- intersect(names(gobp_GEM$genesets), names(gobp_protein$genesets))
gobp_GEM$genesets <- gobp_GEM$genesets[gobp_inter]
gobp_protein$genesets <- gobp_protein$genesets[gobp_inter]

print(length(gobp_inter))
```


```{r}
gomf_GEM <- reduce_gsets(gomf, gene_background_GEM, min_size = 5)
gomf_GEM <- trim_gomf(gomf_GEM)
cat("\n")
gomf_protein <- reduce_gsets(gomf, gene_background_protein, min_size = 5)
gomf_protein <- trim_gomf(gomf_protein)

cat("\n")

gomf_inter <- intersect(names(gomf_GEM$genesets), names(gomf_protein$genesets))
gomf_GEM$genesets <- gomf_GEM$genesets[gomf_inter]
gomf_protein$genesets <- gomf_protein$genesets[gomf_inter]

print(length(gomf_inter))
```


```{r}
max_fdr <- 0.01
## prepare genesets
genesets_GEM <- list(hallmark = hallmark_GEM,
                     kegg = kegg_GEM,
                     reactome = reactome_GEM,
                     gobp = gobp_GEM,
                     gomf = gomf_GEM)

genesets_protein <- list(hallmark = hallmark_protein,
                        kegg = kegg_protein,
                        reactome = reactome_protein,
                        gobp = gobp_protein,
                        gomf = gomf_protein)


backgrounds <- c(length(hallmark_inter), length(kegg_inter), 
                 length(reactome_inter),length(gobp_inter), length(gomf_inter))
```

#### Undirectional 
```{r}
## metabolites w/o background correction
met2gene_signatures <- list(up = table_protein$met2gene_list$up,
                            dn = table_protein$met2gene_list$dn)


## proteins w/o background correction
prot_signatures <- list(up = table_protein$signif_gene_up,
                        dn = table_protein$signif_gene_dn)
df <- data.frame()

for(i in 1:length(genesets_GEM)){
  
  hyp_met <- hypeR(met2gene_signatures, 
                   genesets_GEM[[i]], 
                   test="hypergeometric", 
                   background=length(GEM_gene_map$symbol))
  
  hyp_prot <- hypeR(prot_signatures, 
                    genesets_protein[[i]], 
                    test="hypergeometric", 
                    background=nrow(protein_diffanal))
  
  res <- pair_evaluation_enrichment(hyp_met,
                                   hyp_prot,
                                   background = backgrounds[i],
                                   max_fdr)
  df <- rbind(df, res)
}
colnames(df) <- c("met_enriched","diffanal_enriched","overlap","pval","background", "hit_pathway")
row.names(df) <- c("hallmark", "kegg","reactome","gobp","gomf")


df %>% 
  dplyr::select(-c("hit_pathway")) %>% 
  DT::datatable(.) 

```


#### Directional 
```{r}
## metabolites w/o background correction
met2gene_signatures <- list(up = table_protein_di$met2gene_list$up,
                            dn = table_protein_di$met2gene_list$dn)

## proteins w/o background correction
prot_signatures <- list(up = table_protein_di$signif_gene_up,
                        dn = table_protein_di$signif_gene_dn)


df <- data.frame()
for(i in 1:length(genesets_GEM)){
  
  hyp_met <- hypeR(met2gene_signatures, 
                  genesets_GEM[[i]], 
                  test="hypergeometric", 
                  background=length(Human_GEM_tables$gene_df$name))
  
  hyp_prot <- hypeR(prot_signatures, 
                    genesets_protein[[i]], 
                    test="hypergeometric", 
                    background=nrow(protein_diffanal))
  
  res <- pair_evaluation_enrichment(hyp_met,
                                   hyp_prot,
                                   background = backgrounds[i],
                                   max_fdr)
  df <- rbind(df, res)
}
colnames(df) <- c("met_enriched","diffanal_enriched","overlap","pval","background","hit_pathway")
row.names(df) <- c("hallmark", "kegg","reactome","gobp","gomf")

df %>% 
  dplyr::select(-c("hit_pathway")) %>% 
  DT::datatable(.) 
```

## Enrichment analysis on hypeR-GEM object {.tabset}

### HALLMAKR {.tabset}

#### Unweighted + undirectional
```{r}
uu_hallmark <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj,
                      genesets = hallmark_GEM$genesets,
                      genesets_name = "HALLMARK",
                      method = "unweighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(uu_hallmark, file = file.path(PATH, "data/External/enrichment/urine/uu_hallmark.rds"))
```

#### Weighted + undirectional

```{r}
wu_hallmark <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj,
                        genesets = hallmark_GEM$genesets,
                        genesets_name = "HALLMARK",
                        method = "weighted",
                        min_metabolite = 2,
                        background = length(GEM_gene_map$symbol))

saveRDS(wu_hallmark, file = file.path(PATH, "data/External/enrichment/urine/wu_hallmark.rds"))
```

#### Unweighted + directional

```{r}
ud_hallmark <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj_di,
                          genesets = hallmark_GEM$genesets,
                          genesets_name = "HALLMARK",
                          method = "unweighted",
                          min_metabolite = 2,
                          background = length(GEM_gene_map$symbol))

saveRDS(ud_hallmark, file = file.path(PATH, "data/External/enrichment/urine/ud_hallmark.rds"))
```

#### Weighted + directional

```{r}
wd_hallmark <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj_di,
                          genesets = hallmark_GEM$genesets,
                          genesets_name = "HALLMARK",
                          method = "weighted",
                          min_metabolite = 2,
                          background = length(GEM_gene_map$symbol))

saveRDS(wd_hallmark, file = file.path(PATH, "data/External/enrichment/urine/wd_hallmark.rds"))
```



### KEGG {.tabset}

#### Unweighted + undirectional
```{r}
uu_kegg <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj,
                      genesets = kegg_GEM$genesets,
                      genesets_name = "KEGG",
                      method = "unweighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(uu_kegg, file = file.path(PATH, "data/External/enrichment/urine/uu_kegg.rds"))
```


#### Weighted + undirectional

```{r}
wu_kegg <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj,
                      genesets = kegg_GEM$genesets,
                      genesets_name = "KEGG",
                      method = "weighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(wu_kegg, file = file.path(PATH, "data/External/enrichment/urine/wu_kegg.rds"))
```

#### Unweighted + directional
```{r}
ud_kegg <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj_di,
                      genesets = kegg_GEM$genesets,
                      genesets_name = "KEGG",
                      method = "unweighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(ud_kegg, file = file.path(PATH, "data/External/enrichment/urine/ud_kegg.rds"))
```


#### Weighted + directional
```{r}
wd_kegg <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj_di,
                      genesets = kegg_GEM$genesets,
                      genesets_name = "KEGG",
                      method = "weighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(wd_kegg, file = file.path(PATH, "data/External/enrichment/urine/wd_kegg.rds"))
```


### REACTOME {.tabset}

#### Unweighted + undirectional

```{r}
uu_reactome <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj,
                          genesets = reactome_GEM$genesets,
                          genesets_name = "REACTOME",
                          method = "unweighted",
                          min_metabolite = 2,
                          background = length(GEM_gene_map$symbol))

saveRDS(uu_reactome, file = file.path(PATH, "data/External/enrichment/urine/uu_reactome.rds"))
```


#### Weighted + undirectional

```{r}
wu_reactome <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj,
                      genesets = reactome_GEM$genesets,
                      genesets_name = "REACTOME",
                      method = "weighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(wu_reactome, file = file.path(PATH, "data/External/enrichment/urine/wu_reactome.rds"))
```


#### Unweighted + directional
```{r}
ud_reactome <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj_di,
                      genesets = reactome_GEM$genesets,
                      genesets_name = "REACTOME",
                      method = "unweighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(ud_reactome, file = file.path(PATH, "data/External/enrichment/urine/ud_reactome.rds"))
```


#### Weighted + directional
```{r}
wd_reactome <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj_di,
                      genesets = reactome_GEM$genesets,
                      genesets_name = "REACTOME",
                      method = "weighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(wd_reactome, file = file.path(PATH, "data/External/enrichment/urine/wd_reactome.rds"))
```

### GOBP {.tabset}

#### Unweighted + undirectional

```{r}
uu_gobp <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj,
                      genesets = gobp_GEM$genesets,
                      genesets_name = "GOBP",
                      method = "unweighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(uu_gobp, file = file.path(PATH, "data/External/enrichment/urine/uu_gobp.rds"))
```


#### Weighted + undirectional

```{r}
wu_gobp <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj,
                      genesets = gobp_GEM$genesets,
                      genesets_name = "GOBP",
                      method = "weighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(wu_gobp, file = file.path(PATH, "data/External/enrichment/urine/wu_gobp.rds"))
```

#### Unweighted + directional
```{r}
ud_gobp <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj_di,
                      genesets = gobp_GEM$genesets,
                      genesets_name = "GOBP",
                      method = "unweighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(ud_gobp, file = file.path(PATH, "data/External/enrichment/urine/ud_gobp.rds"))
```


#### Weighted + directional
```{r}
wd_gobp <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj_di,
                      genesets = gobp_GEM$genesets,
                      genesets_name = "GOBP",
                      method = "weighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(wd_gobp, file = file.path(PATH, "data/External/enrichment/urine/wd_gobp.rds"))
```

### GOMF {.tabset}

#### Unweighted + undirectional

```{r}
uu_gomf <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj,
                      genesets = gomf_GEM$genesets,
                      genesets_name = "GOMF",
                      method = "unweighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(uu_gomf, file = file.path(PATH, "data/External/enrichment/urine/uu_gomf.rds"))
```


#### Weighted + undirectional

```{r}
wu_gomf <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj,
                      genesets = gomf_GEM$genesets,
                      genesets_name = "GOMF",
                      method = "weighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(wu_gomf, file = file.path(PATH, "data/External/enrichment/urine/wu_gomf.rds"))
```

#### Unweighted + directional
```{r}
ud_gomf <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj_di,
                      genesets = gomf_GEM$genesets,
                      genesets_name = "GOMF",
                      method = "unweighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(ud_gomf, file = file.path(PATH, "data/External/enrichment/urine/ud_gomf.rds"))
```


#### Weighted + directional
```{r}
wd_gomf <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj_di,
                      genesets = gomf_GEM$genesets,
                      genesets_name = "GOMF",
                      method = "weighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(wd_gomf, file = file.path(PATH, "data/External/enrichment/urine/wd_gomf.rds"))
```



