---
title: "M005 Plasma differential analysis using linear model"
author: "Ziwei Huang"
date:   "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: united
    code_folding: hide
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = FALSE,warning = FALSE)
```

## Setting
```{r}
rm(list=ls())
suppressPackageStartupMessages(library(Biobase))
suppressPackageStartupMessages(library(GSVA))
suppressPackageStartupMessages(library(Matrix))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lme4))
suppressPackageStartupMessages(library(lmerTest))
suppressPackageStartupMessages(library(hypeR))
suppressPackageStartupMessages(library(gprofiler2))
suppressPackageStartupMessages(library(DT))
suppressPackageStartupMessages(library(SummarizedExperiment))

PATH <- file.path(Sys.getenv("NECS"),"Ziwei_Analysis","hypeR.GEM_evaluation")
devtools::load_all(file.path(PATH, "../in_dev_packages/hypeR.GEM"))

source(file.path(PATH,"R/utility.R"))
```

## Load preprocessed data

- C12m (control) vs. Aca (Acarbose) (x)

- C12m (control) vs. 17aE2 (17-a-E2) (x)

- C12m (control) vs. Cana (Canaglif) (x)

- C12m (control) vs. CR (CR) (?)

- C12m (control) vs. Rapa (Rapa) (?)

- C12m (control) vs. Young (Y4m) (x)

- GHRWT (GHR wild type) vs. GHRKO (GHR KO) (x)

- SnellWT (Snell wile type) vs. SnellDW (Snell DW) (good)

- covariates = Sex or Sex + Age (some missing values in Age)

```{r}
## Metabolomics
SumExp_met <- readRDS(file = file.path(PATH, "data/M005/metabolomics/plasma_wmissing.rds"))

pdata <- colData(SumExp_met) %>% 
  as.data.frame(.) %>% 
  dplyr::rename_with(~ gsub("\\.x$", "", .), ends_with(".x")) %>% 
  dplyr::select(-ends_with(".y")) %>% 
  dplyr::mutate(Sex = as.factor(Sex))

fdata <- rowData(SumExp_met) %>% 
  as.data.frame(.) %>% 
  dplyr::rename_with(~ gsub("\\.x$", "", .), ends_with(".x")) %>% 
  dplyr::select(-ends_with(".y"))

SumExp_met <- SummarizedExperiment(assays = assay(SumExp_met),
                                   colData = pdata,
                                   rowData = fdata)

## Use metabolites labeled by refmet_name
SumExp_met <- SumExp_met[!is.na(rowData(SumExp_met)$refmet_name), ]


## Proteomics
SumExp_prot <- readRDS(file = file.path(PATH, "data/M005/proteomics/plasma_wmissing.rds"))

pdata <- colData(SumExp_prot) %>% 
  as.data.frame(.) %>% 
  dplyr::rename_with(~ gsub("\\.x$", "", .), ends_with(".x")) %>% 
  dplyr::select(-ends_with(".y")) %>% 
  dplyr::mutate(Sex = as.factor(Sex))

SumExp_prot <- SummarizedExperiment(assays = assay(SumExp_prot),
                                    colData = pdata)

uniprot_mapping <- readRDS(file = file.path(PATH, "data/M005/proteomics/plasma_protein_uniprot_mapping.rds"))
```

## Data Distribution

### Proteomics

#### Expression
```{r}
hist(assay(SumExp_prot),
     main="Plasma Proteomics")
```

#### Expression per sample
```{r}
boxplot(assay(SumExp_prot),
        main="Plasma Proteomics")
```


### Metabolomics

#### Expression
```{r}
hist(assay(SumExp_met),
     main="Plasma Metabolomics")
```

#### Expression per sample
```{r}
boxplot(assay(SumExp_met),
        main="Plasma Metabolomics")
```


## SnellWT vs. SnellDW {.tabset}

### Subset and imputation
```{r}
missing_threshold_met <- 0.4
missing_threshold_prot <- 0.3

## Subset
pheno_pair <- c("SnellWT", "SnellDW")

SumExp_met <- SumExp_met[, colData(SumExp_met)$Intervention %in% pheno_pair]
colData(SumExp_met)$Intervention <- factor(colData(SumExp_met)$Intervention, levels = pheno_pair)

SumExp_prot <- SumExp_prot[, colData(SumExp_prot)$Intervention %in% pheno_pair]
colData(SumExp_prot)$Intervention <- factor(colData(SumExp_prot)$Intervention, levels = pheno_pair)


## Imputation
missing_ratio_per_met <- apply(assay(SumExp_met),1,function(x){return (sum(is.na(x))/length(x))})
SumExp_met <- SumExp_met[missing_ratio_per_met <= missing_threshold_met, ]
assay(SumExp_met) <- halfmin_impute(assay(SumExp_met))

missing_ratio_per_protein <- apply(assay(SumExp_prot),1,function(x){return (sum(is.na(x))/length(x))})
SumExp_prot <- SumExp_prot[missing_ratio_per_protein <= missing_threshold_prot, ]
assay(SumExp_prot) <- halfmin_impute(assay(SumExp_prot))

data.frame() %>% 
  rbind(c(dim(SumExp_met)[1], dim(SumExp_met)[2])) %>% 
  rbind(c(dim(SumExp_prot)[1], dim(SumExp_prot)[2])) %>% 
  magrittr::set_colnames(c("feature", "sample")) %>% 
  magrittr::set_rownames(c("metabolomics", "proteomics")) %>% 
  DT::datatable(.)
```

### Metabolomics
```{r}
run <- FALSE
max_fdr <-  0.05

if(run){
  met_diffanal <- apply( assay(SumExp_met), 1, function(y){
        tmp <- lm(y ~ Intervention + Sex, data = colData(SumExp_met))
        return(summary(tmp)$coefficients["InterventionSnellDW",c("Estimate","Std. Error","t value","Pr(>|t|)")])
        }) %>% 
      t(.) %>%
      data.frame(.) %>%
      dplyr::rename(estimate="Estimate",se="Std..Error", tvalue = "t.value", pvalue="Pr...t..") %>%
      dplyr::mutate(fdr=p.adjust(pvalue,method="BH")) %>%
      tibble::rownames_to_column(var="symbol") %>%
      dplyr::mutate(gene_type  = case_when(
        estimate > 0 & fdr < max_fdr ~ "up",
        estimate < 0 & fdr < max_fdr ~ "dn",
        TRUE ~ "ns"
      ))

  refmet <- as.data.frame(rowData(SumExp_met)) %>% 
    tibble::rownames_to_column(var="symbol") %>% 
    dplyr::select(c("symbol","refmet_name"))
  
  met_diffanal <- met_diffanal %>% 
    dplyr::left_join(refmet, by="symbol") %>% 
    dplyr::select(c("symbol","refmet_name"), everything())
  
  saveRDS(met_diffanal, file = file.path(PATH, "data/M005/metabolomics/plasma_met_diffanal.rds"))
}else{
  met_diffanal <- readRDS(file.path(PATH, "data/M005/metabolomics/plasma_met_diffanal.rds"))
}
met_up <- met_diffanal %>% 
  dplyr::filter(gene_type == "up")

met_dn <- met_diffanal %>% 
  dplyr::filter(gene_type == "dn")

met_sigs <- list(up = met_up,
                  dn = met_dn)

## check
sapply(met_sigs ,nrow) %>%  tibble::enframe(value = "size")
```

### Proteomics

```{r}
run <- FALSE
max_fdr <-  0.01

if(run){
  protein_diffanal <- apply( assay(SumExp_prot), 1, function(y){
        tmp <- lm(y ~ Intervention + Sex, data = colData(SumExp_prot))
        return(summary(tmp)$coefficients["InterventionSnellDW",c("Estimate","Std. Error","t value","Pr(>|t|)")])
        }) %>% 
      t(.) %>%
      data.frame(.) %>%
      dplyr::rename(estimate="Estimate",se="Std..Error", tvalue = "t.value", pvalue="Pr...t..")
      
  
  protein_diffanal <- protein_diffanal %>% 
    tibble::rownames_to_column(var="symbol") %>% 
    dplyr::rename(uni_prot_symbol = symbol) %>% 
    dplyr::left_join(uniprot_mapping, by="uni_prot_symbol") %>% 
    dplyr::mutate(fdr=p.adjust(pvalue,method="BH")) %>% 
    dplyr::mutate(gene_type  = case_when(
        estimate > 0 & fdr < max_fdr ~ "up",
        estimate < 0 & fdr < max_fdr ~ "dn",
        TRUE ~ "ns"
      )) %>% 
    #dplyr::select(-c("uni_prot_symbol","protein_names")) %>% 
    dplyr::select(symbol, everything())
  
  saveRDS(protein_diffanal, file = file.path(PATH, "data/M005/proteomics/plasma_protein_diffanal.rds"))
}else{
  protein_diffanal <- readRDS(file = file.path(PATH, "data/M005/proteomics/plasma_protein_diffanal.rds"))
}

prot_up <- protein_diffanal %>%
  dplyr::filter(gene_type == "up")

prot_dn <- protein_diffanal %>%
  dplyr::filter(gene_type == "dn")

prot_sigs <- list(up = prot_up,
                  dn = prot_dn)

## check
sapply(prot_sigs ,nrow) %>%  tibble::enframe(value = "size")
```


### GEM mapping {.tabset}

```{r}
data("Human_GEM_tables")
GEM_gene_map <- gprofiler2::gconvert(query = Human_GEM_tables$gene_df$name,
                                     organism = "hsapiens",
                                     target = "ENSG",
                                     filter_na = TRUE) %>%
   dplyr::select(input, name) %>%
   dplyr::rename(ensemble_ID = input, symbol = name)

```

#### Undirectional mapping
```{r}
hypeR_GEM_obj <- hypeR.GEM::signature2gene(signatures = met_sigs,
                                           species = "Human",
                                           directional = FALSE,
                                           merge = TRUE,
                                           promiscuous_threshold = 300,
                                           ensemble_id = TRUE,
                                           reference_key = 'refmet_name',
                                           background = NULL)

## check
sapply(hypeR_GEM_obj$gene_tables,nrow) %>%  tibble::enframe(value = "size")

saveRDS(hypeR_GEM_obj, file = file.path(PATH, "data/M005/plasma_hypR_GEM_object.rds"))
```

### Directional mapping
```{r}
hypeR_GEM_obj_di <- hypeR.GEM::signature2gene(signatures = met_sigs,
                                             species = "Human",
                                             directional = TRUE,
                                             merge = TRUE,
                                             promiscuous_threshold = 300,
                                             ensemble_id = TRUE,
                                             reference_key = 'refmet_name',
                                             background = NULL)

## check
sapply(hypeR_GEM_obj_di$gene_tables,nrow) %>%  tibble::enframe(value = "size")

saveRDS(hypeR_GEM_obj_di, file = file.path(PATH, "data/M005/plasma_hypR_GEM_object_di.rds"))
```


## Pair evaluation (gene-level, hypergeometric) {.tabset}

- Take overlap between hypeR-GEM mapped genes and genes in proteomics ($GEM \cap genes\ in\ proteomics$)

- Counter how many of these overlap genes are significant

- Background = ($total GEM \cap proteomics$)

- setA = ($GEM\ mapped \cap Background$)

- setB = ($significant\ genes\ in\ proteomics \cap Background$)

```{r}
tmp <- sapply(met_sigs ,nrow) %>%  tibble::enframe(value = "size")
background_list <- intersect(protein_diffanal$symbol, GEM_gene_map$symbol)

table_protein <- pair_evaluation(hypeR_GEM_obj,
                                 protein_diffanal,
                                 background_list) 


table_protein$df_grouped <- table_protein$df_grouped %>% 
  dplyr::mutate(DE_met = sum(tmp$size)) %>% 
  dplyr::select(DE_met, everything())

table_protein_di <- pair_evaluation(hypeR_GEM_obj_di,
                                   protein_diffanal,
                                   background_list) 


table_protein_di$df_grouped <- table_protein_di$df_grouped %>% 
  dplyr::mutate(DE_met = sum(tmp$size)) %>% 
  dplyr::select(DE_met, everything())


saveRDS(table_protein, file = file.path(PATH, "data/M005/plasma_table_protein.rds"))
saveRDS(table_protein_di, file = file.path(PATH, "data/M005/plasma_table_protein_di.rds"))
```

### Undirectional vs. Protein_diffanal
```{r}
table_protein$df_grouped %>%  DT::datatable(.)
```


### Directional vs. Protein_diffanal
```{r}
table_protein_di$df_grouped %>%  DT::datatable(.)
```

## Pair evaluation (pathway-level, hypergeometric)

### Load genesets
```{r}
hallmark <- hypeR::msigdb_gsets(species="Homo sapiens", "H",clean = TRUE)
kegg <- hypeR::msigdb_gsets(species="Homo sapiens", "C2", "CP:KEGG_LEGACY", clean = TRUE)
reactome <- hyperdb_rgsets("REACTOME", "70.0")
gobp <- hypeR::msigdb_gsets(species="Homo sapiens", "C5", "BP", clean = TRUE)
gomf <- hypeR::msigdb_gsets(species="Homo sapiens", "C5", "MF", clean = TRUE)

## background of each omics
gene_background_GEM <- GEM_gene_map$symbol
gene_background_protein <- protein_diffanal$symbol
```


```{r}
hallmark_GEM <- reduce_gsets(hallmark, gene_background_GEM, min_size = 5)
cat("\n")
hallmark_protein <- reduce_gsets(hallmark, gene_background_protein, min_size = 5)

cat("\n")

hallmark_inter <- intersect(names(hallmark_GEM$genesets), names(hallmark_protein$genesets))
hallmark_GEM$genesets <- hallmark_GEM$genesets[hallmark_inter]
hallmark_protein$genesets <- hallmark_protein$genesets[hallmark_inter]

print(length(hallmark_inter))
```


```{r}
kegg_GEM <- reduce_gsets(kegg, gene_background_GEM, min_size = 5)
cat("\n")
kegg_protein <- reduce_gsets(kegg, gene_background_protein, min_size = 5)

cat("\n")

kegg_inter <- intersect(names(kegg_GEM$genesets), names(kegg_protein$genesets))
kegg_GEM$genesets <- kegg_GEM$genesets[kegg_inter]
kegg_protein$genesets <- kegg_protein$genesets[kegg_inter]

print(length(kegg_inter))
```


```{r}
reactome_GEM <- reduce_gsets(reactome, gene_background_GEM, min_size = 5)
cat("\n")
reactome_protein <- reduce_gsets(reactome, gene_background_protein, min_size = 5)

cat("\n")

reactome_inter <- intersect(names(reactome_GEM$genesets), names(reactome_protein$genesets))
reactome_GEM$genesets <- reactome_GEM$genesets[reactome_inter]
reactome_protein$genesets <- reactome_protein$genesets[reactome_inter]

print(length(reactome_inter))
```


```{r}
gobp_GEM <- reduce_gsets(gobp, gene_background_GEM, min_size = 5)
cat("\n")
gobp_protein <- reduce_gsets(gobp, gene_background_protein, min_size = 5)

cat("\n")

gobp_inter <- intersect(names(gobp_GEM$genesets), names(gobp_protein$genesets))
gobp_GEM$genesets <- gobp_GEM$genesets[gobp_inter]
gobp_protein$genesets <- gobp_protein$genesets[gobp_inter]

print(length(gobp_inter))
```

```{r}
gomf_GEM <- reduce_gsets(gomf, gene_background_GEM, min_size = 5)
gomf_GEM <- trim_gomf(gomf_GEM)
cat("\n")
gomf_protein <- reduce_gsets(gomf, gene_background_protein, min_size = 5)
gomf_protein <- trim_gomf(gomf_protein)

cat("\n")

gomf_inter <- intersect(names(gomf_GEM$genesets), names(gomf_protein$genesets))
gomf_GEM$genesets <- gomf_GEM$genesets[gomf_inter]
gomf_protein$genesets <- gomf_protein$genesets[gomf_inter]

print(length(gomf_inter))
```

```{r}
max_fdr <- 0.01
## prepare genesets
genesets_GEM <- list(hallmark = hallmark_GEM,
                     kegg = kegg_GEM,
                     reactome = reactome_GEM,
                     gobp = gobp_GEM,
                     gomf = gomf_GEM)

genesets_protein <- list(hallmark = hallmark_protein,
                        kegg = kegg_protein,
                        reactome = reactome_protein,
                        gobp = gobp_protein,
                        gomf = gomf_protein)


backgrounds <- c(length(hallmark_inter), length(kegg_inter), 
                 length(reactome_inter),length(gobp_inter), length(gomf_inter))
```

#### Undirectional 
```{r}
## metabolites w/o background correction
met2gene_signatures <- list(up = table_protein$met2gene_list$up,
                            dn = table_protein$met2gene_list$dn)


## proteins w/o background correction
prot_signatures <- list(up = table_protein$signif_gene_up,
                        dn = table_protein$signif_gene_dn)
df <- data.frame()

for(i in 1:length(genesets_GEM)){
  
  hyp_met <- hypeR(met2gene_signatures, 
                   genesets_GEM[[i]], 
                   test="hypergeometric", 
                   background=length(GEM_gene_map$symbol))
  
  hyp_prot <- hypeR(prot_signatures, 
                    genesets_protein[[i]], 
                    test="hypergeometric", 
                    background=nrow(protein_diffanal))
  
  res <- pair_evaluation_enrichment(hyp_met,
                                   hyp_prot,
                                   background = backgrounds[i],
                                   max_fdr)
  df <- rbind(df, res)
}
colnames(df) <- c("met_enriched","diffanal_enriched","overlap","pval","background", "hit_pathway")
row.names(df) <- c("hallmark", "kegg","reactome","gobp","gomf")


df %>% 
  dplyr::select(-c("hit_pathway")) %>% 
  DT::datatable(.) 
```

#### Directional 
```{r}
## metabolites w/o background correction
met2gene_signatures <- list(up = table_protein_di$met2gene_list$up,
                            dn = table_protein_di$met2gene_list$dn)

## proteins w/o background correction
prot_signatures <- list(up = table_protein_di$signif_gene_up,
                        dn = table_protein_di$signif_gene_dn)


df <- data.frame()
for(i in 1:length(genesets_GEM)){
  
  hyp_met <- hypeR(met2gene_signatures, 
                  genesets_GEM[[i]], 
                  test="hypergeometric", 
                  background=length(Human_GEM_tables$gene_df$name))
  
  hyp_prot <- hypeR(prot_signatures, 
                    genesets_protein[[i]], 
                    test="hypergeometric", 
                    background=nrow(protein_diffanal))
  
  res <- pair_evaluation_enrichment(hyp_met,
                                   hyp_prot,
                                   background = backgrounds[i],
                                   max_fdr)
  df <- rbind(df, res)
}
colnames(df) <- c("met_enriched","diffanal_enriched","overlap","pval","background","hit_pathway")
row.names(df) <- c("hallmark", "kegg","reactome","gobp","gomf")

df %>% 
  dplyr::select(-c("hit_pathway")) %>% 
  DT::datatable(.) 
```

## Enrichment analysis on hypeR-GEM object {.tabset}

```{r}
#max_fdr <- 0.01
```

### HALLMAKR {.tabset}

#### Unweighted + undirectional
```{r}
uu_hallmark <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj,
                      genesets = hallmark_GEM$genesets,
                      genesets_name = "HALLMARK",
                      method = "unweighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(uu_hallmark, file = file.path(PATH, "data/M005/plasma_enrichment/uu_hallmark.rds"))
```

#### Weighted + undirectional

```{r}
wu_hallmark <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj,
                        genesets = hallmark_GEM$genesets,
                        genesets_name = "HALLMARK",
                        method = "weighted",
                        min_metabolite = 2,
                        background = length(GEM_gene_map$symbol))

saveRDS(wu_hallmark, file = file.path(PATH, "data/M005/plasma_enrichment/wu_hallmark.rds"))
```

#### Unweighted + directional

```{r}
ud_hallmark <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj_di,
                          genesets = hallmark_GEM$genesets,
                          genesets_name = "HALLMARK",
                          method = "unweighted",
                          min_metabolite = 2,
                          background = length(GEM_gene_map$symbol))

saveRDS(ud_hallmark, file = file.path(PATH, "data/M005/plasma_enrichment/ud_hallmark.rds"))
```

#### Weighted + directional

```{r}
wd_hallmark <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj_di,
                          genesets = hallmark_GEM$genesets,
                          genesets_name = "HALLMARK",
                          method = "weighted",
                          min_metabolite = 2,
                          background = length(GEM_gene_map$symbol))

saveRDS(wd_hallmark, file = file.path(PATH, "data/M005/plasma_enrichment/wd_hallmark.rds"))
```



### KEGG {.tabset}

#### Unweighted + undirectional
```{r}
uu_kegg <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj,
                      genesets = kegg_GEM$genesets,
                      genesets_name = "KEGG",
                      method = "unweighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(uu_kegg, file = file.path(PATH, "data/M005/plasma_enrichment/uu_kegg.rds"))
```


#### Weighted + undirectional

```{r}
wu_kegg <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj,
                      genesets = kegg_GEM$genesets,
                      genesets_name = "KEGG",
                      method = "weighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(wu_kegg, file = file.path(PATH, "data/M005/plasma_enrichment/wu_kegg.rds"))
```

#### Unweighted + directional
```{r}
ud_kegg <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj_di,
                      genesets = kegg_GEM$genesets,
                      genesets_name = "KEGG",
                      method = "unweighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(ud_kegg, file = file.path(PATH, "data/M005/plasma_enrichment/ud_kegg.rds"))
```


#### Weighted + directional
```{r}
wd_kegg <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj_di,
                      genesets = kegg_GEM$genesets,
                      genesets_name = "KEGG",
                      method = "weighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(wd_kegg, file = file.path(PATH, "data/M005/plasma_enrichment/wd_kegg.rds"))
```


### REACTOME {.tabset}

#### Unweighted + undirectional

```{r}
uu_reactome <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj,
                          genesets = reactome_GEM$genesets,
                          genesets_name = "REACTOME",
                          method = "unweighted",
                          min_metabolite = 2,
                          background = length(GEM_gene_map$symbol))

saveRDS(uu_reactome, file = file.path(PATH, "data/M005/plasma_enrichment/uu_reactome.rds"))
```


#### Weighted + undirectional

```{r}
wu_reactome <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj,
                      genesets = reactome_GEM$genesets,
                      genesets_name = "REACTOME",
                      method = "weighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(wu_reactome, file = file.path(PATH, "data/M005/plasma_enrichment/wu_reactome.rds"))
```


#### Unweighted + directional
```{r}
ud_reactome <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj_di,
                      genesets = reactome_GEM$genesets,
                      genesets_name = "REACTOME",
                      method = "unweighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(ud_reactome, file = file.path(PATH, "data/M005/plasma_enrichment/ud_reactome.rds"))
```


#### Weighted + directional
```{r}
wd_reactome <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj_di,
                      genesets = reactome_GEM$genesets,
                      genesets_name = "REACTOME",
                      method = "weighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(wd_reactome, file = file.path(PATH, "data/M005/plasma_enrichment/wd_reactome.rds"))
```

### GOBP {.tabset}

#### Unweighted + undirectional

```{r}
uu_gobp <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj,
                      genesets = gobp_GEM$genesets,
                      genesets_name = "GOBP",
                      method = "unweighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(uu_gobp, file = file.path(PATH, "data/M005/plasma_enrichment/uu_gobp.rds"))
```


#### Weighted + undirectional

```{r}
wu_gobp <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj,
                      genesets = gobp_GEM$genesets,
                      genesets_name = "GOBP",
                      method = "weighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(wu_gobp, file = file.path(PATH, "data/M005/plasma_enrichment/wu_gobp.rds"))
```

#### Unweighted + directional
```{r}
ud_gobp <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj_di,
                      genesets = gobp_GEM$genesets,
                      genesets_name = "GOBP",
                      method = "unweighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(ud_gobp, file = file.path(PATH, "data/M005/plasma_enrichment/ud_gobp.rds"))
```


#### Weighted + directional
```{r}
wd_gobp <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj_di,
                      genesets = gobp_GEM$genesets,
                      genesets_name = "GOBP",
                      method = "weighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(wd_gobp, file = file.path(PATH, "data/M005/plasma_enrichment/wd_gobp.rds"))
```

### GOMF {.tabset}

#### Unweighted + undirectional

```{r}
uu_gomf <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj,
                      genesets = gomf_GEM$genesets,
                      genesets_name = "GOMF",
                      method = "unweighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(uu_gomf, file = file.path(PATH, "data/M005/plasma_enrichment/uu_gomf.rds"))
```


#### Weighted + undirectional

```{r}
wu_gomf <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj,
                      genesets = gomf_GEM$genesets,
                      genesets_name = "GOMF",
                      method = "weighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(wu_gomf, file = file.path(PATH, "data/M005/plasma_enrichment/wu_gomf.rds"))
```

#### Unweighted + directional
```{r}
ud_gomf <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj_di,
                      genesets = gomf_GEM$genesets,
                      genesets_name = "GOMF",
                      method = "unweighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(ud_gomf, file = file.path(PATH, "data/M005/plasma_enrichment/ud_gomf.rds"))
```


#### Weighted + directional
```{r}
wd_gomf <- enrichment(hypeR_GEM_obj = hypeR_GEM_obj_di,
                      genesets = gomf_GEM$genesets,
                      genesets_name = "GOMF",
                      method = "weighted",
                      min_metabolite = 2,
                      background = length(GEM_gene_map$symbol))

saveRDS(wd_gomf, file = file.path(PATH, "data/M005/plasma_enrichment/wd_gomf.rds"))
```

## Metabolomics enrichment

### Load sets
```{r}
## MSETS
#M_metabolon_hmdbID <- readRDS(file = file.path(PATH, "data/MSETS/MSETS_metabolon_hmdbID.rds"))
M_metabolon_refmetID <- readRDS(file = file.path(PATH, "data/MSETS/MSETS_metabolon_refmetID.rds"))

## RMSETS
R_refmet_refmetID <- readRDS(file = file.path(PATH, "data/MSETS/RMSETS_refmet_necs2022_refmetID_v1.0.rds"))
R_metabolon_refmetID <- readRDS(file = file.path(PATH, "data/MSETS/RMSETS_metabolon_necs2022_refmetID_v1.0.rds"))
                                
## Background
refmet_bg <- met_diffanal %>% dplyr::filter(!is.na(refmet_name)) %>% dplyr::pull("refmet_name")
```


### Prepare metabolic signatures

```{r}
met_diffanal_up_refmetID <- met_diffanal  %>% 
  dplyr::filter(gene_type == "up") %>% 
  dplyr::filter(!is.na(refmet_name)) %>% 
  dplyr::pull("refmet_name")

met_diffanal_dn_refmetID <- met_diffanal  %>% 
  dplyr::filter(gene_type == "dn") %>% 
  dplyr::filter(!is.na(refmet_name)) %>% 
  dplyr::pull("refmet_name")

met_diffanal_refmetID <- list(up = met_diffanal_up_refmetID,
                              dn =met_diffanal_dn_refmetID)
```


### MSETS enrichment {.tabset}

```{r}
mset_reduce <- function(mset,
                        background,
                        min_size){
  
  mset_flt <- lapply(mset, intersect, background)
  mset_flt <- mset_flt[lengths(mset_flt) >= min_size]
  
  return(mset_flt)
}
```

```{r}
min_size <- 5

## metabolonSupPthwy
M_metabolonSupPthwy_flt <- mset_reduce(M_metabolon_refmetID$metabolonSupPthwy, refmet_bg, min_size)
  
## metabolonSubPthwy
M_metabolonSubPthwy_flt <- mset_reduce(M_metabolon_refmetID$metabolonSubPthwy, refmet_bg, min_size)

## refmetSuperClass
refmetSuperClass_flt <- mset_reduce(M_metabolon_refmetID$refmetSuperClass, refmet_bg, min_size)

## refmetMainClass
refmetMainClass_flt <- mset_reduce(M_metabolon_refmetID$refmetMainClass, refmet_bg, min_size)

## refmetSubClass
refmetSubClass_flt <- mset_reduce(M_metabolon_refmetID$refmetSubClass, refmet_bg, min_size)
```

#### metabolonSupPthwy
```{r}
hyp_metabolonSupPthwy <- hypeR(met_diffanal_refmetID , 
                               M_metabolonSupPthwy_flt, 
                               test="hypergeometric", 
                               background=nrow(met_diffanal))

saveRDS(hyp_metabolonSupPthwy, file = file.path(PATH, "data/M005/plasma_enrichment/Met_M_hyp_metabolonSupPthwy.rds"))
```

#### metabolonSubPthwy
```{r}
hyp_metabolonSubPthwy <- hypeR(met_diffanal_refmetID , 
                               M_metabolonSubPthwy_flt, 
                               test="hypergeometric", 
                               background=nrow(met_diffanal))

saveRDS(hyp_metabolonSubPthwy, file = file.path(PATH, "data/M005/plasma_enrichment/Met_M_hyp_metabolonSubPthwy.rds"))
```

#### refmetSuperClass

```{r}
hyp_refmetSuperClass <- hypeR(met_diffanal_refmetID , 
                              refmetSuperClass_flt, 
                              test="hypergeometric", 
                              background=nrow(met_diffanal))

saveRDS(hyp_refmetSuperClass, file = file.path(PATH, "data/M005/plasma_enrichment/Met_M_hyp_refmetSuperClass.rds"))
```

#### refmetMainClass

```{r}
hyp_refmetMainClass <- hypeR(met_diffanal_refmetID , 
                             refmetMainClass_flt, 
                             test="hypergeometric", 
                             background=nrow(met_diffanal))

saveRDS(hyp_refmetMainClass, file = file.path(PATH, "data/M005/plasma_enrichment/Met_M_hyp_refmetMainClass.rds"))
```

#### refmetSubClass
```{r}
hyp_refmetSubClass <- hypeR(met_diffanal_refmetID , 
                             refmetSubClass_flt, 
                             test="hypergeometric", 
                             background=nrow(met_diffanal))

saveRDS(hyp_refmetSubClass, file = file.path(PATH, "data/M005/plasma_enrichment/Met_M_hyp_refmetSubClass.rds"))
```

### RMSETS enrichment {.tabset}

```{r}
R_metabolon_flt <- mset_reduce(R_metabolon_refmetID$genesets, refmet_bg, min_size)
R_refmet_flt <- mset_reduce(R_refmet_refmetID$genesets, refmet_bg, min_size)
```

#### Metabolon
```{r}
hyp_metabolon <- hypeR(met_diffanal_refmetID , 
                       R_metabolon_flt, 
                       test="hypergeometric", 
                       background=nrow(met_diffanal))


saveRDS(hyp_metabolon, file = file.path(PATH, "data/M005/plasma_enrichment/Met_R_hyp_metabolon.rds"))
```

#### Refmet
```{r}
hyp_refmet <- hypeR(met_diffanal_refmetID , 
                       R_refmet_flt, 
                       test="hypergeometric", 
                       background=nrow(met_diffanal))

saveRDS(hyp_refmet, file = file.path(PATH, "data/M005/plasma_enrichment/Met_R_hyp_refmet.rds"))
```
